---
title: "Additional seaweed samples (validation)"
output: html_notebook
---

Let's import the additional samples to check whether there were differences prior to antibiotic TimepointCat.

```{r}
library(mbtools)

ps <- readRDS("data/additional_seaweed.rds")
sdata <- sample_data(ps) %>% as("data.frame")
sdata[["Subject"]] <- substr(sdata$id, 1, 2)
sdata[["Days"]] <- substr(sdata$id, 4, 5) %>% as.numeric()
manifest <- readRDS("data/seaweed_exp.rds") %>% sample_data() %>% as("data.frame")
sdata <- merge(
  sdata, 
  unique(manifest[, c("Subject", "Response")]), 
  by = "Subject"
)
sdata[["TimepointCat"]] <- c("before", "during")[(sdata$Days >= 27) + 1] %>% factor()
rownames(sdata) <- sdata$group_id
sample_data(ps) <- sdata
ps
```

Now lets have a look at ordination.

```{r, fig.width=6, fig.height=4}
theme_set(theme_bw())

rare <- rarefy_even_depth(ps, 20000)
sample_data(rare)$Days <- factor(sample_data(rare)$Days)
ord <- ordinate(rare, method="PCoA")

data <- plot_ordination(rare, ord, color="Response", shape="TimepointCat", label="Subject", justDF=T)
plot_ordination(rare, ord, color="Response", shape="TimepointCat", label="Subject") +
    stat_ellipse(data = data[data$TimepointCat == "before", ],
                 aes(group = TimepointCat), type = "t", lty="dashed") +
    stat_ellipse(aes(group = Response), type = "t") +
    scale_shape_manual(values = c(before = 1, during = 10)) +
    labs(shape = "sample time")
ggsave("figures/additional_pcoa.svg", width=6, height=4)
```

Let's see if the response ype can not be distinguished from the initial samples.


```{r}
before <- subset_samples(rare, TimepointCat == "before")
model <- vegan::adonis(otu_table(before) ~ Response, data=as(sample_data(before), "data.frame"))
model
```

So can not be distinguished before, but can be distinguished after:

```{r}
during <- subset_samples(rare, TimepointCat == "during")
model <- vegan::adonis(otu_table(during) ~ Response, data=as(sample_data(during), "data.frame"))
model
```

```{r}
phyla <- taxa_count(ps, "Phylum")

sdata <- as.data.table(as(sample_data(ps), "data.frame"))
phyla <- phyla[sdata, on = c(sample = "group_id")]
head(phyla)
```

Phyla distributions:

```{r, fig.height = 5, fig.width = 8}
phyla[, relative := reads / sum(reads), by = "sample"]
phyla[, "seaweed" := "seaweed diet"]
phyla[grepl("A|B", Subject), seaweed := "normal diet"]

high_taxa <- phyla[relative > 1e-2, unique(taxa)] 

pl <- ggplot(phyla[taxa %in% high_taxa], 
             aes(x = factor(Days), y = relative, fill = taxa)) + 
  geom_bar(stat = "identity", position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  facet_grid(seaweed ~ Response) + 
  theme_classic() +
  labs(x = "day", y = "relative abundance", fill = "phylum")
ggsave("figures/phyla_day_seaweed.png", height = 5, width = 8, dpi = 300)
pl
```

## Differential tests

We will use corn cob to test for differences in abundance between responders and non-responders for each timepoint category.

We will use a few helpers to summarize the tables and corncob output.

```{r}
source("helpers.R")
```

### For phyla

```{r}
library(corncob)

phylum <- subset_samples(ps, Response != "control") %>% collapse("Phylum")
cats <- levels(sample_data(phylum)$TimepointCat)
xtabs(~ Response + TimepointCat, as(sample_data(phylum), "data.frame"))

tests <- lapply(cats, function(cat) {
  tests <- differentialTest(
    formula = ~ Response,
    phi.formula = ~ 1,
    formula_null = ~ 1,
    phi.formula_null = ~ 1,
    data = prune_samples(sample_data(phylum)$TimepointCat == cat, phylum),
    test = "LRT",
    boot = F
  ) %>% results_table()
  tests$TimepointCat <- cat
  tests
}) %>% rbindlist()
```

```{r}
fwrite(tests[order(padj)], "data/tests_additional_phyla.csv")
tests[padj<0.05, table(TimepointCat)]
tests[padj < 0.05]
```

For genera:

```{r}
genera <- subset_samples(ps, Response != "control") %>% collapse("Genus")
cats <- levels(sample_data(genera)$TimepointCat)

tests <- lapply(cats, function(cat) {
  tests <- differentialTest(
    formula = ~ Response,
    phi.formula = ~ 1,
    formula_null = ~ 1,
    phi.formula_null = ~ 1,
    data = prune_samples(sample_data(genera)$TimepointCat == cat, genera),
    test = "LRT",
    boot = F
  ) %>% results_table()
  tests$TimepointCat <- cat
  tests
}) %>% rbindlist()

```

```{r}
fwrite(tests[order(padj)], "data/tests_additional_genera.csv")
tests[padj<0.05, table(TimepointCat)]
tests[padj < 0.05 & TimepointCat == "before"]
```