---
title: "Antibiotics quantification"
output: html_notebook
---

# Antibiotics quantification

```{r}
library(mbtools)
theme_set(theme_bw())
```

## Dilution curve

```{r}
dilution <- fread("data/seaweed/cefoperazone_calibration_2ul.csv")
dilution[, c("id", "replicate") := tstrsplit(Replicate, "_")]
dilution[, "concentration_on_column" := `concentration ug/mL` / 8]
mod <- lm(concentration_on_column ~ 0 + Area, data = dilution)
summary(mod)
ggplot(dilution, aes(x=Area, y=concentration_on_column)) + geom_point() + stat_smooth(method="lm")
```

## Fecal samples

Let's start by reading the data.

```{r}
cefoperazone_fecal <- fread("data/seaweed/cefoperazone_fecal_5ul.csv")
cefoperazone_fecal[, 
  c("hplc_batch", "id", "replicate") := tstrsplit(Replicate, "_")]
cefoperazone_fecal[, c("mouse", "day") := list(
  substr(id, 1, 2), substr(id, 4, 5))]
cefoperazone_fecal[, "Replicate" := NULL]
metadata_fecal <- fread("data/seaweed/cefoperazone_fecal_metadata.csv")[
  , "day" := as.character(day)]
cefoperazone_fecal <- metadata_fecal[cefoperazone_fecal, on = c("id", "day")]
```

Now let's compare the levels of cefoperazone. Men fecal pellet weight was 50mg which was suspended in a volume of 850ul. This olution was then diluted 1:250 before loading. We also injected 2ul in the calibration curve but 5ul here. So we will adjust for that as well.

```{r, fig.width=8, fig.height=3}
library(ggplot2)
theme_set(theme_minimal())

cef <- cefoperazone_fecal[Compound == "Cefoperazone" & `Fragment Ion` == "Cefoperazone_t1"]
cef[, "concentration" := coef(mod)[1] * Area * 2/5 * 850/50 * 250]
ggplot(cef, aes(x = day, y = concentration, color = response)) +
  geom_hline(yintercept=4, lty="dashed") +
  geom_boxplot(outlier.color = NA) + geom_jitter(height = 0, width = 0.2) +
  labs(y = "fecal concentration [µg/mL]", x="") + guides(color=F) +
  scale_y_log10() +
  facet_wrap(~ response)
ggsave("figures/fecal_AB.svg", width=8, height=3)
```


## Plasma data

For blood we dilute 50ul plasma in a total volume of 600ul and this was diluted 1:1 again after centrifugation. So we have a total dilution of 24-fold.

```{r}
cefoperazone_plasma <- fread("data/seaweed/cefoperazone_plasma_5ul.csv")
cefoperazone_plasma[, 
  c("mouse", "replicate") := tstrsplit(Replicate, "_")]
cefoperazone_plasma[, "Replicate" := NULL]
metadata_plasma <- fread("data/seaweed/cefoperazone_plasma_metadata.csv")[
  , "day" := as.character(day)][, "mouse" := id]
cefoperazone_plasma <- metadata_plasma[cefoperazone_plasma, on = c("mouse")]
```

And let's visualize this again:

```{r, fig.width = 4, fig.height=3}
cef <- cefoperazone_plasma[Compound == "Cefoperazone" & `Fragment Ion` == "Cefoperazone_t1"]
cef[, "concentration" := coef(mod)[1] * Area * 2/5 * 24 * 1e3]
ggplot(cef, aes(x = response, y = concentration, color = response)) +
  labs(y = "plasma concentration [ng/mL]", x="") +
  geom_boxplot() + geom_jitter(height = 0, width = 0.2) +
  scale_y_log10()
ggsave("figures/blood_AB.svg", width=4, height=3)
```