---
title: "R Notebook"
author: "Christian Diener <cdiener@systemsbiology.org>"
output: html_notebook
---

# Effect of duration length

## Data preparation

We will start by reading our data and required libraries.

```{r}
library(mbtools)
theme_set(theme_bw())

ps <- readRDS("data/duration_exp.rds")

total <- subset_samples(ps, Spore == 0)
```

We sill start by building up a dataframe that combines phyla abundances and sample metadata.

```{r}
phyla <- taxa_count(total, "Phylum")

sdata <- as.data.table(as(sample_data(total), "data.frame"))
phyla <- phyla[sdata, on = c(sample = "Description")]
head(phyla)
```


# Phyla abundances

## By duration

First we have to assemble the relative abundances for controls.

```{r}
control <- phyla[duration == 0]
control[, relative := reads / sum(reads, na.rm = T)]

high_taxa <- phyla[Day - duration == 0, 
                   .(taxa, rel = reads / sum(reads)), 
                   by = "sample"][rel > 1e-3, unique(taxa)] 
```

We will plot a reference bar chart for those, removing all phyla that contribute less than 0.1% of the total reads.

```{r, fig.height=4, fig.width=1.5}
pl <- ggplot(control[taxa %in% high_taxa], 
             aes(x = factor(duration), y = relative, fill = taxa)) + 
  geom_bar(stat = "identity") + scale_y_continuous(labels = scales::percent) +
  theme_bw() + 
  labs(x = "antibiotic duration [days]", y = "relative abundance") +
  facet_wrap(~ Response) + guides(fill = "none") + labs(x = "")
ggsave("figures/control_duration.png", height = 4, width = 1.5, dpi = 300)
pl
```

Now we will do the same thing for the non-responders and responders.

```{r, fig.height=4, fig.width=8}
treated <- phyla[duration > 0 & Day - duration == 0]
treated[, relative := reads / sum(reads), by = c("duration", "Response")]

pl <- ggplot(treated[taxa %in% high_taxa], 
             aes(x = factor(duration), y = relative, fill = taxa)) + 
  geom_bar(stat = "identity") + scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ Response) + theme_bw() +
  labs(x = "", y = "", fill = "phylum")
ggsave("figures/treatment_duration.png", height = 4, width = 8, dpi = 300)
pl
```

# Community collap

We can see that most of the Cyanobacteria and Proteobacteria are actually Chloroplasts and Mitochondria:

```{r}
non_responders <- subset_samples(total, Response == "responder")
cyano_proteo <- subset_taxa(non_responders, 
                            Phylum %in% c("Cyanobacteria", "Proteobacteria"))
class <- taxa_count(cyano_proteo, "Order")[order(-reads), sum(reads), by = taxa]
fam <- taxa_count(cyano_proteo, "Family")[order(-reads), sum(reads), by = taxa]
print(class[, .(taxa, relative = V1 / sum(V1))])
print(fam[, .(taxa, relative = V1 / sum(V1))])
```

Microbial collapse usually implies a lower bacterial biomass compared with background signals such as DNA from food and the host.

```{r}
chloro_mit <- subset_taxa(ps, 
                          Class == "Chloroplast" | Family == "Mitochondria")
rel <- data.table(sample = sample_names(ps), 
                  relative = sample_sums(chloro_mit) / sample_sums(ps))
rel <- sdata[rel, on = c(Description = "sample")]
```

Let's have a look at the proportion of Mitochondria or Choloroplasts by duration.

```{r, fig.width=3, fig.height=4}
ggplot(rel, aes(x = TimepointCat, y = relative, color = Response)) +
  geom_hline(yintercept = 0.1, lty="dashed") +
  geom_jitter(width=0.2) +
  labs(x="sample time", y="Mitochondria/Chloroplasts [fraction of total]") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
ggsave("figures/duration_chloro_mit.svg", width=3, height=4)
```

Corresponding test

```{r}
print(rel[TimepointCat == "during", mean(relative), by="Response"])[, diff(V1)]
wilcox.test(relative ~ Response, rel[TimepointCat == "during"])
```

## Summarized

```{r, fig.width=4, fig.height=5}
library(gridExtra)

treated <- phyla[duration >= 0 & Day - duration == 0]
treated[, relative := reads / sum(reads), by = "Response"]

bars <- ggplot(treated[taxa %in% high_taxa], 
               aes(x = Response, y = relative, fill = taxa)) + 
  geom_bar(stat = "identity") + scale_y_continuous(labels = scales::percent) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)) +
  labs(x = "", y = "relative abundance", fill = "phylum")
after_duration <- sdata[duration > 0 & Day - duration == 0]
resp_counts <- after_duration[, table(Response, duration)][2:3, ]
tbl <- tableGrob(resp_counts)
pl <- grid.arrange(bars, tbl, nrow = 2, as.table = T, heights = c(3, 1))
ggsave("figures/summary_duration.svg", plot = pl, height = 6, width = 4)
pl
```

And the corresponding Fisher test:

```{r}
fisher.test(resp_counts) %>% print
```

# Beta diversity

We are interested in the beta diversity between samples after treatment, so we will start by selecting a subset of those
and calculating the ordination. We will look at the sequencing depth to get a decent depth for rarefaction.

```{r}
treat <- subset_samples(total, Day >= 0)
sample_sums(treat)
```

Okay so we can easily rarefy to 10K reads.

```{r}
ord <- rarefy_even_depth(treat, 10000) %>% 
       ordinate(., method = "PCoA", distance = "bray")
sda <- sample_data(treat)
sample_data(treat)$since_treatment <- factor(sda$Day - sda$duration)

data <- plot_ordination(treat, ord, color = "Response", 
                      shape = "TimepointCat", justDF = T)
pl <- plot_ordination(treat, ord, color = "Response", 
                      shape = "TimepointCat") +
      stat_ellipse(data=data[data$TimepointCat == "before", ], 
                   aes(group = TimepointCat), lty="dashed") +
      stat_ellipse(aes(group = Response), type = "t") +
      scale_shape_manual(values = c(before = 1, during = 10, after = 19)) +
      labs(shape = "sample time") + theme_bw()

ggsave("figures/treatment_pcoa_duration.svg", width = 6, height = 4)
pl  
```

# Corresponding PERMANOVA

Before exposure:
```{r}
library(vegan)

ps <- subset_samples(treat, TimepointCat == "before") %>% 
      rarefy_even_depth(., 10000)
counts <- as.matrix(taxa_count(ps, NA))
sda <- as(sample_data(ps), "data.frame")
permanova <- adonis(counts ~ Response, data = sda)
permanova
```

After exposure:
```{r}
library(vegan)

ps <- subset_samples(treat, TimepointCat != "before") %>% 
      rarefy_even_depth(., 10000)
counts <- as.matrix(taxa_count(ps, NA))
sda <- as(sample_data(ps), "data.frame")
permanova <- adonis(counts ~ Response, data = sda)
permanova
```

# Gained/lost taxa

We will now track which taxa where gained or lost in each mouse over the treatment course.

```{r}
asvs <- taxa_count(total, NA)
asvs <- asvs[sdata, on = c(sample = "Description")][Day %in% c(0, 32)]
asvs
```

Now lets us assign which ASVs have persisted, were gained or lost.

```{r}
status <- function(dt) {
  if (nrow(dt) == 2) return("persisted")
  if (nrow(dt) == 1 && dt$Day == 0) return("lost")
  return("gained")
}

history <- asvs[reads > 0, .(status = status(.SD), taxon = taxa[1], 
                             mouse = Mouse[1], duration = duration[1], 
                             response = Response[1]), 
                by = c("Mouse", "taxa")]
status_counts <- history[, .(n = .N, duration = duration[1], 
                             response = response[1]), 
                         by = c("mouse", "status")]
status_counts[, duration := factor(duration)]
```

Now let's plot all of that:

```{r, fig.height=3, fig.width=8}
pl <- ggplot(status_counts, 
             aes(x = response, y = n, group = duration, color = duration)) +
  geom_boxplot(col = "gray", aes(group = response), outlier.shape = NA) +
  geom_point(position = position_dodge(width = 0.6)) + 
  facet_wrap(~ status) + theme_bw() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)) +
  labs(x = "", y = "ASVs [count]", color = "duration [days]")
ggsave("figures/avs_history.svg", height = 3, width = 8)
pl
```

And the corresponding tests:

control vs. non-responder

```{r, message=FALSE}
for (lev in status_counts[, unique(status)]) {
  cat(lev)
  pairwise.wilcox.test(status_counts[status == lev, n], 
                       status_counts[status == lev, response], 
                       p.adjust.method = "bonferroni") %>% print
}
```

# Time courses

## Shannon index

```{r, fig.width=5, fig.height=4}
shannon <- estimate_richness(rarefy_even_depth(total, 10000), 
                             measures = "Shannon")
shannon <- data.table(sample = rownames(shannon), shannon = shannon[, 1])
shannon <- shannon[sdata, on = c(sample = "Description")]
shannon[Day >= 0 & Day < 32, Day := Day - duration]
shannon[, "exposure" := sprintf("%dd exposure", duration)]
shannon[duration == 0, exposure := "no antibiotics"]

pl <- ggplot(shannon[Day >= -1 & Day < 10], 
             aes(x = Day, y = shannon, color = Response)) + 
  geom_point(aes(group=Mouse)) +
  geom_line(aes(group=Mouse)) +  
  #stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), 
  #             geom = "pointrange") +
  #stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "line") +
  xlim(-1, 8) + theme_bw() + 
  labs(y = "alpha diversity [Shannon index]", x = "days after treatment", color = "") +
  facet_wrap(~ reorder(exposure, duration)) +
  theme(legend.position = c(0.85, 0.25))
ggsave("figures/shannon_time.svg", width = 6, height = 4)
pl
```

## Differential tests

We will use corn cob to test for differences in abundance between responders and non-responders for each timepoint category.

We will use a few helpers to summarize the tables and corncob output.

```{r}
source("helpers.R")
```

### For phyla

Controls vs non-responders

```{r}
library(corncob)

phylum <- subset_samples(treat, TimepointCat == "before") %>% collapse("Phylum", 1e-3)

tests <- differentialTest(
  formula = ~ Response,
  phi.formula = ~ 1,
  formula_null = ~ 1,
  phi.formula_null = ~ 1,
  data = phylum,
  test = "LRT",
  boot = F
) %>% results_table()

fwrite(tests, "data/tests_duration_before_phyla.csv")
tests[order(padj)]
```

Non-responders vs. responders

```{r}
library(corncob)

phylum <- subset_samples(treat, Response != "control") %>% collapse("Phylum")
cats <- levels(sample_data(phylum)$TimepointCat)
xtabs(~ Response + TimepointCat, as(sample_data(phylum), "data.frame"))

tests <- lapply(cats, function(cat) {
  tests <- differentialTest(
    formula = ~ Response,
    phi.formula = ~ 1,
    formula_null = ~ 1,
    phi.formula_null = ~ 1,
    data = prune_samples(sample_data(phylum)$TimepointCat == cat, phylum),
    test = "LRT",
    boot = F
  ) %>% results_table()
  tests$TimepointCat <- cat
  tests
}) %>% rbindlist()
```

```{r}
fwrite(tests[order(padj)], "data/tests_duration_during_phyla.csv")
tests[padj<0.05, table(TimepointCat)]
tests[padj < 0.05]
```

For genera:

```{r}
library(corncob)

genera <- subset_samples(treat, TimepointCat == "before") %>% collapse("Genus", 1e-3, 0.8)

tests <- differentialTest(
  formula = ~ Response,
  phi.formula = ~ 1,
  formula_null = ~ 1,
  phi.formula_null = ~ 1,
  data = genera,
  test = "LRT",
  boot = F
) %>% results_table()

fwrite(tests, "data/tests_duration_before_genera.csv")
tests[order(padj)]
```

```{r}
genera <- subset_samples(treat, Response != "control") %>% collapse("Genus")
cats <- levels(sample_data(genera)$TimepointCat)

tests <- lapply(cats, function(cat) {
  tests <- differentialTest(
    formula = ~ Response,
    phi.formula = ~ 1,
    formula_null = ~ 1,
    phi.formula_null = ~ 1,
    data = prune_samples(sample_data(genera)$TimepointCat == cat, genera),
    test = "LRT",
    boot = F
  ) %>% results_table()
  tests$TimepointCat <- cat
  tests
}) %>% rbindlist()

```

```{r}
fwrite(tests[order(padj)], "data/tests_duration_during_genera.csv")
tests[padj<0.05, table(TimepointCat)]
tests[order(padj)]
```

## Major phyla timecourses

```{r, fig.height = 4, fig.width = 6}
shifted <- phyla[Day - duration >= 0 & Day <= 32]
shifted[, Day := Day - duration]
shifted[, relative := reads / sum(reads), by = "sample"]

pl <- ggplot(shifted[taxa %in% c("Bacteroidetes", "Firmicutes")], 
             aes(x = Day, y = relative, col = taxa)) +
  geom_point(alpha = 0.5) + scale_y_continuous(labels = scales::percent) +
  geom_line(aes(group = interaction(Mouse, taxa)), alpha = 0.5) + 
  facet_wrap(~ Response, nrow = 3, strip.position = "right") + theme_bw() +
  labs(x = "days after treatment", y = "relative abundance")
ggsave("figures/bac_firm_time.svg", width = 6, height = 4)
pl
```