---
title: "Preprocessing of the data"
author: "Christian Diener <cdiener@systemsbioogy.org>"
output: html_notebook
---

# Converting to a phyloseq object

# duration experiment

We will start by reading the DADA2 results and metadata and bundling everything
in a phyloseq object.

```{r}
library(phyloseq)
library(stringr)

dada <- readRDS("data/duration/dada2.rds")
meta <- read.table("data/duration/metadata.tsv", 
                   comment.char = "#", header = 1)
rownames(meta) <- meta$Description
meta <- meta[rownames(dada$feature_table), ]
duration <- phyloseq(otu_table(dada$feature_table, taxa_are_rows = FALSE),
                     tax_table(dada$taxonomy),
                     sample_data(meta))
```

We will extract the duration durations from the sample names.

```{r}
map <- c(cnt = "control", non = "non-responder", res = "responder")

sdata <- sample_data(duration)
sdata$duration <- str_replace(sdata$Mouse, "[A-Z]", "") %>% as.numeric()
sdata$TimepointCat <- "before"
sdata$TimepointCat[sdata$Day == sdata$duration] <- "during"
sdata$TimepointCat[sdata$Day > sdata$duration] <- "after"
sdata$TimepointCat[sdata$Response == "cnt"] <- "before"
sdata$TimepointCat <- factor(sdata$TimepointCat, 
                             levels = c("before", "during", "after"))
sdata$Response <- factor(map[as.character(sdata$Response)], levels = map)
sample_data(duration) <- sdata
```

We will save the phyloseq object for further analyses.

```{r}
saveRDS(duration, "data/duration_exp.rds")
```

# Seaweed experiment

Again we start by adding the metadata and bundling as a phyloseq object.

```{r}
dada <- readRDS("data/seaweed/dada2.rds")
meta <- read.table("data/seaweed/metadata.tsv", 
                   comment.char = "#", header = 1)
rownames(meta) <- tolower(meta$CategoricalID)
meta <- meta[rownames(dada$feature_table), ]
seaweed <- phyloseq(otu_table(dada$feature_table, taxa_are_rows = FALSE),
                    tax_table(dada$taxonomy),
                    sample_data(meta))
```

Groups A and B received no seaweed in their diet whereas groups C and D did. We will
also remove 4 samples where the diet was switched right before antibiotics (internal controls).

```{r}
sample_data(seaweed)$seaweed <- sample_data(seaweed)$Group %in% c("C", "D")
sample_data(seaweed)$CategoricalID <- tolower(sample_data(seaweed)$CategoricalID)
seaweed <- subset_samples(seaweed, !(Subject %in% c("A9", "A10", "C9", 
                                                    "C10", "X1", "X2", "Y1", 
                                                    "Y2")))
```

Let's clean up annotations and add some variables:

```{r}
sdata <- as(sample_data(seaweed), "data.frame")
sdata$TimepointCat <- "before"
sdata$TimepointCat[sdata$Days == 31] <- "during"
sdata$TimepointCat[sdata$Days > 31] <- "after"
sdata$TimepointCat[sdata$Response == "cnt"] <- "before"
sdata$TimepointCat <- factor(sdata$TimepointCat, 
                             levels = c("before", "during", "after"))
sdata$Response <- factor(map[as.character(sdata$Response)], 
                         levels = map)
sample_data(seaweed) <- sdata
```

We will also merge in the qPCR cycle time (Cq or Ct values):

```{r}
Ct <- read.table("data/seaweed/qpcr_data.tsv", header = TRUE)
names(Ct) <- c("CategoricalID", "Ct")
merged <- merge(sdata, Ct, by="CategoricalID")
rownames(merged) <- merged$CategoricalID
sample_data(seaweed) <- merged
```

We will save the phyloseq object for further analyses.

```{r}
saveRDS(seaweed, "data/seaweed_exp.rds")
```