---
title: "16S Power Analysis"
output: html_notebook
---

In order to perform power analysis for 16S data we will need a decent distribution model 
for the data. We will model counts using Beta-Binomial model as used in 
[corncob](https://arxiv.org/pdf/1902.02776.pdf). For that you will need to install the
`corncob` package first.

```{r}
devtools::install_github("bryandmartin/corncob")
```

We can now start running power analyses for the data. For that we will use a reference dataset
generated in the same lab and under the same protocol as our datasets (described in
https://doi.org/10.1016/j.celrep.2018.07.032). We will only use control samples.

```{r}
library(mbtools)

seqtab <- readRDS(url(
  "https://data.mendeley.com/public-files/datasets/86xzh5c85p/files/e547053f-094a-46a5-9e7b-e0d3867b698c/file_downloaded"))
taxa <- readRDS(url(
  "https://data.mendeley.com/public-files/datasets/86xzh5c85p/files/df0f29dd-f66f-48e5-b3af-1ee4f3ffbf2c/file_downloaded"))
meta <- read.csv(url(
  "https://data.mendeley.com/public-files/datasets/86xzh5c85p/files/294bf92b-5484-4cbe-94bc-81cf290a25d3/file_downloaded"))
rownames(meta) <- rownames(seqtab)
ref <- phyloseq(otu_table(seqtab, taxa_are_rows = FALSE), tax_table(taxa),
                sample_data(meta))
ref <- subset_samples(ref, Treatment == "NC")
```

This leaves us with 48 samples. We will now generate new data from the fitted
Beta-Binomial of the reference data. Many ASVs will be omitted in that step
since they have zero counts in all of the control samples.

# PERMANOVA power analysis

We will start by testing the power of PERMANOVA. Power will bes estimated across 100
sampled count matrices with 10 replicates (1000 samples in total). The effect size used
here is the inverse fold change. Meaning if we use an effect of $\alpha$ for an ASV with proportion
$p_i$ the treated group will be assigned a proportion $(1 - \alpha)\cdot p_i$. If we would do this
for all ASVs we would obtain a mean Bray-Curtis distance of $\alpha$ between any control and treatment
sample. If only half of all ASVs are differential this would yield a mean distance of $\alpha/2$ and
so on.

```{r}
if (!file.exists("data/permanova_power.csv")) {
  pa <- power_analysis(ref, fraction_differential = 0.1,
                       n = ceiling(2 ^ seq(2, 6, length.out = 12) / 2) * 2)
  power <- pa$power
  fwrite(power, "data/permanova_power.csv")
} else {
  power <- fread("data/permanova_power.csv")
}
```

We can now plot the power curves for the different $R^2$ values:

```{r, fig.width = 5, fig.height = 3}
theme_set(theme_minimal())

ggplot(power[effect > 0], aes(x = n, y = power, col = asym_r2, 
                              group = effect)) +
  scale_x_log10() + 
  geom_hline(yintercept = 0.8, lty = "dashed") + 
  geom_point(size = 2) + geom_line(size = 1) +
  geom_linerange(aes(ymin = power - power_sd, ymax = power + power_sd)) +
  labs(color = expression(R^2))
ggsave("figures/permanova_power.svg", width = 5, height = 3)
```

We can also look at the false discovery rate which is simply the power for a zero effect (no true difference).

```{r, fig.width = 4, fig.height = 3}
ggplot(power[effect == 0], aes(x = n, y = power, group = effect)) +
  scale_x_log10() + ylim(0, 0.1) +
  geom_hline(yintercept = 0.05, lty = "dashed") + 
  geom_point() + geom_line() +
  geom_linerange(aes(ymin = power - power_sd, ymax = power + power_sd)) +
  labs(y = "false discovery rate", color = "R2 [log10]")
ggsave("figures/permanova_fdr.svg", width = 4, height = 3)
```

# Mann-Whitney Power Analysis

We can do the same for the Mann-Whitney test on proportions. This is run for each taxa. We will never
look at ranks more detailed than genus so we will summarize the data on that level. One thing to note
is that power will depend on the proportion. Very rare taxa will have lower power.

```{r}
if (!file.exists("data/mw_power.csv")) {
  pa <- power_analysis(tax_glom(ref, "Genus"), 
                       fraction_differential = 0.5, method = "mw")
  power <- pa$power
  fwrite(power, "data/mw_power.csv")
} else {
  mw_power <- fread("data/mw_power.csv")
}
```

And we can visualize the power curves again. Since we have an additional level of information
this time (individual taxa), we will stratify by the effect size (fold change).

```{r, fig.width = 5, fig.height = 3}
power <- mw_power[mu>1e-3, .(power = mean(power), power_sd = sd(power)), by = c("effect", "n")]
power[, "fold_change" := 1 / (1 - effect)]
label <- as_labeller(function(x) sprintf("fold change = %.2gx", as.numeric(x)))
ggplot(power[effect > 0], aes(x = n, y = power, col=factor(round(fold_change, 1)), group=fold_change)) +
  scale_x_log10() +
  geom_hline(yintercept = 0.8, lty = "dashed") + 
  geom_point(size=2) + geom_line(size = 1) +
  geom_linerange(aes(ymin = power - power_sd, ymax = power + power_sd),
                 alpha = 0.5) +
  labs(color = "fold change")
ggsave("figures/mw_power.png", width = 5, height = 3, dpi = 300)
```

And we can agin estimate the FDR:

```{r, fig.width = 4, fig.height = 3}
ggplot(power[effect == 0], aes(x = n, y = power)) +
  scale_x_log10() + ylim(0, 0.1) +
  geom_hline(yintercept = 0.05, lty = "dashed") + 
  geom_point() + geom_line() +
  geom_linerange(aes(ymin = power - power_sd, ymax = power + power_sd)) + 
  labs(y = "false discovery rate", color = "signal/noise [log10]")
ggsave("figures/mw_fdr.svg", width = 4, height = 3)
```